Land-use legacies and climate change as a double challenge to oak forest resilience: mismatches of geographical and ecological rear edges

### 2020-08-05 man/ecosystems/ms_ecosystems.Rmd
update version 3

The response of species to changing environments (*e.g.* distributional shifts) can be largely determined by population responses at range margins [@HampePetit2005ConservingBiodiversity]. Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range [*i.e.* center-periphery hypothesis; @SagarinGaines2002AbundantCentre; @Pirononetal2017GeographicVariation]. It has been often assumed that geographically marginal populations represent ecological marginal populations. This means lower performance, higher vulnerability, and thus higher risk of extinction than for populations at the core of the species' range [@Rehmetal2015LosingYour; @Pirononetal2017GeographicVariation; @VilaCabreraetal2019RefiningPredictions]. Nonetheless, recent reviews report that species- and population-specific responses do not always support this hypothesis [@Sextonetal2009EvolutionEcology; @Abelietal2014EffectsMarginality; @Oldfatheretal2020RangeEdges]. This is partly explained because rear-edge is a multidimensional concept including an ecological (*i.e.* climatic and edaphic), a geographical and a genetic component [@VilaCabreraetal2019RefiningPredictions], but also an anthropogenic (*i.e.* land-use) dimension. In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, it is crucial to identify and understand mismatches between the geographical and the ecological rear-edges [@VilaCabreraJump2019GreaterGrowth].

Limits of species distribution are strongly determined by climatic factors and biotic interactions [@Gaston2009GeographicRange; @Sextonetal2009EvolutionEcology]. Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and there are already signs that more intense and longer droughts are altering forest dynamics [@Allenetal2010GlobalOverview]. Drought frequency and severity have increased in recent decades, with a trend towards drier summers, particularly for Southern Europe [@VicenteSerranoetal2014EvidenceIncreasing; @Staggeetal2017ObservedDrought]. In this climatic-change context, population loss and range retractions are expected in boreal, temperate, and Mediterranean species at the lowest latitudes, elevations, and drought-prone areas of a species' distribution, *i.e.* the rear edge. The rear-edge populations are likely to be more sensitive to minor climatic and microtopographic variations and therefore the effects of droughts are expected to be particularly relevant [@HampePetit2005ConservingBiodiversity; @VilaCabreraetal2019RefiningPredictions].

It is often overlooked that anthropic-actions are a driver of change as relevant or more as natural drivers like climate [@Salaetal2000GlobalBiodiversity], particularly for regions with long land-use history like the Mediterranean Region [*e.g.* @NavarroGonzalezetal2013WeightLanduse; @DoblasMirandaetal2017ReviewCombination]. In these areas, the susceptibility and response of ecosystems to natural disturbances are conditioned by legacies of historical land-use activity [*e.g.* @Munteanuetal2015Legacies19th; @Mausolfetal2018LegacyEffects]. The past land-use legacies interact with recent natural disturbances and may confound their interpretation [@Fosteretal2003ImportanceLanduse]. For example, recent works showed that a quarter of current forests in the Iberian Peninsula, the rear edge of several temperate and boreal tree species, are growing on former agricultural and grazing land abandoned after the 1950s [@VilaCabreraetal2017NewForests]. Consequently, anthropogenic habitat modification and its legacies represent a critical dimension of marginality as they may intensify, confound or delay climate driven population decline at rear edges [@VilaCabreraJump2019GreaterGrowth; @SanchezdeDiosetal2020FagusSylvatica]. In this context, the main focus of our work is to identify the impacts and responses to natural (severe drought) and anthropogenic (*e.g.* logging) disturbances on oak forests at their southern geographical range. Thus, the incorporation of an historical perspective should aid to better interpret the responses of ecosystems to natural disturbances [@Fosteretal2003ImportanceLanduse], and particularly that of marginal rear edge populations [@VilaCabreraetal2019RefiningPredictions].

The assessment of resilience to climate and human disturbances provides critical information concerning the capacity of forests to maintain their structure and render valuable ecosystem services. Resilience is the capacity of an ecosystem to persist and maintain its state and functions in the face of disturbance [@Holling1973ResilienceStability; @Hodgsonetal2015WhatYou]. @Lloretetal2011ComponentsTree proposed an approach, which decomposes resilience to drought in three components: resistance to drought, recovery after drought and resilience. Forest resilience is determined by the capacity to reduce the impact (resistance) and the capacity to recover from the impact of disturbance (recovery) [@IngrischBahn2018ComparableQuantification]. This conceptual approach has been very popular to assess forest resilience recently, because it allows a simple, yet highly efficient assessment of short-term responses of trees to drought. Nevertheless, it is not exempted from some criticism and needs to be applied carefully to avoid over interpretation and potential bias associated at different levels [@Schwarzetal2020QuantifyingGrowth]. In this sense we assessed forest resilience both in the long-term to several recent extreme drought episodes and to climate change (*i.e.* warming) in the long-term (the last few decades) using two different proxies to characterize resilience. Dendroecological estimates of growth (*i.e.* tree-ring width) are commonly used proxies to characterize tree vitality, and, generally annual tree-ring widths have been used to study growth changes in response to drought at the individual tree level [@Fritts1976TreeRings; @Dobbertin2005TreeGrowth]. Remote sensing can be used to analyze the impact of drought on ecosystems and hence at the stand level [*e.g.* @Zhangetal2013MonitoringEstimating]. The combination of remote sensing and dendroecology has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerranoetal2013ResponseVegetation; @Coulthardetal2017TreeGrowth], and also to evaluate growth resilience to drought in several tree species [*e.g.* @Gazoletal2018ForestResilience; @PenaGallardoetal2018DroughtSensitiveness].

In the present study, we assessed resilience of *Quercus pyrenaica* Willd. (Pyrenean oak) from southern relict forests at the rear-edge of the species distribution, where species performance is considered to be threatened by climate change [@GeaIzquierdoetal2013GrowthProjections; @GeaIzquierdoetal2017RiskyFuture]. We combined remote-sensing information and dendroecological methods to evaluate the impact of drought both in canopy greenness (as a proxy for primary growth) and radial tree growth (as a proxy for secondary growth). The analysis of resilience to climate was discussed taken into account the land-use history of these transformed forests. To do so we conducted a review of historical documents to reconstruct forest history in our study sites, and to determine how anthropogenic drivers have shaped the current forest structure. Based on this discussion of forest resilience, we specifically developed a rationale that integrates the ecological and anthropogenic dimensions of marginality to determine the regional and local scale mechanisms shaping the probability of persistence (or extinction) of oak rear edge populations. Our main hypothesis is that these stands will show low resilience to extreme droughts along a small-scale gradient. To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests at their present geographical rear edge; (*ii*) conducted and explored a review of historical documents to reconstruct forest management history and to infer how it impacted tree-growth in time; (*iii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iv*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale. For the latter, we characterized variability in the forest response to drought within the current geographical rear-edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience, and recovery to extreme drought events.

@@

The forests of this species reach their southernmost European limit in Andalusian mountains such as Sierra Nevada (37°N, 3°W), a high-mountain range with elevations of up to 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought and increasing aridity with decreasing altitude, and marked variability according to elevation and aspect. Sierra Nevada is considered a glacial refuge for deciduous *Quercus* species [@Olaldeetal2002WhiteOaks]. Eight Pyrenean oak patches (2400 ha) have been identified in this mountain range (Figure 1), from 1100 to 2000 m *a.s.l.* and often associated with major river valleys. Today, *Quercus pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@HampePetit2005ConservingBiodiversity]. They are the richest forest formation in vascular plant species of Sierra Nevada, containing several endemic and endangered plant species [@Loriteetal2008PhytosociologicalReview]. They also harbor high levels of intraspecific genetic diversity [@ValbuenaCarabanaGil2013GeneticResilience]. These relict forests have undergone intensive human use throughout history [@CamachoOlmedoetal2002DinamicaEvolutiva]. Furthermore, the conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer from climate change, reducing its suitable habitats in the near future [@GeaIzquierdoetal2013GrowthProjections; @GeaIzquierdoetal2017RiskyFuture].

@@

The severe drought events since 1901 were identified using SPEI-12. They were characterized in terms of duration, severity and intensity [see @Spinonietal2015EuropeanDrought]. In a first step, the impact of drought in greenness and growth was explored using the $EVI_{sa}$ and the mean RWI site chronologies (Figure 2). Additionally the relationships between climatic variables and tree-growth variables (RWI and BAI site chronologies) were assessed using bootstrapped Pearson's correlations estimated using treeclim [@ZangBiondi2015TreeclimPackage]. The non-climatic disturbance impacts on tree-growth were evaluated using site disturbance chronologies (buit using growth changes on tree growth).

@@

## Results
### Time trends in vegetation greenness
The analysis of time trends in greenness showed that 78.9% of the EVI pixels followed a positive trend for the 2000-2016 period. The lowest values of EVI standardized anomalies for the study period were recorded during the 2005 drought, and the minimum EVI values were expressed in the northern (dry) population (Figure 3a). A "browning" episode ($\mathrm{EVI_{sa}} < -1$) was found during this drought event, whereas no changes in greenness in response to the 2012 drought were detected (Figure 3b).

### Analysis of radial-growth trends and disturbances 
The trees of the southern population were older than those from the northern one. In addition, trees from the southern population at high elevation were taller and their growth was significantly greater than that of trees from the other two sites, despite stand competition measured as plot basal area was greatest in CA-High (Table 1, Figure 4a). The growth and height of trees from the northern and the low-elevation southern population were similar (Figures 4a and S3a). Only trees from the southern sites (*i.e.* the wetter exposure) showed significant positive growth trends since the late 1970s (Figure 4a), this trend being far more pronounced for the wetter and colder high elevation site (CA-High).

@@

### Resilience to drought events at the forest stand and individual-tree levels
Resilience and resistance varied in the same direction whereas recovery varied inversely to resilience and resistance. During the last two drought events, resilience metrics for greenness and tree growth significantly differed between drought events (Table S1). The 2005 drought event reduced greenness and growth more than that of 2012 (Table S2) but the metrics of resilience generally covaried in the same direction during those two years. For EVI, resilience and resistance values were significantly higher for 2012, the most severe event, than for 2005 (Table S2; Figure 5b); whereas recovery values were higher for 2005 than for the 2012 drought event. For BAI, the resilience, resistance and recovery values were higher for 2012 than for 2005 (Table S2, Figure 5c).

@@

## Discussion
By using a combined approach of remote-sensing information and dendroecology, we have quantified growth of adult trees and greenness (EVI) as proxies for secondary and primary growth of relict Mediterranean *Quercus pyrenaica* populations in the southern Iberian Peninsula. These relict oak populations driven by historical land-use have been resilient to climate change at their present rear edge. However, resistance, resilience, and forest recovery to extreme drought events were strongly influenced by mountain exposure, local environmental conditions, and management legacies. This means that the geographical and the ecological rear edges do not necessarily match and, at a small spatial scale, tree performance can vary markedly along the rear edge under climate change.

### High sensitivity and variability in the oak sensitivity to climate at the rear-edge
Severe drought negatively affects both primary and secondary growth of *Q. pyrenaica* forests. This was expressed by the observed reduction in greenness and tree growth in response to the 2005 and 2012 drought events as well as by radial-growth suppression during extreme drought events [@Corcueraetal2006RadialgrowthWoodanatomical; @GeaIzquierdoCanellas2014LocalClimate]. Furthermore, the greatest reduction of tree growth was detected during the 1995 drought, a characteristic negative precipitation anomaly that caused severe and extensive damage in the Mediterranean Iberian Peninsula [@Penuelasetal2001SevereDrought; @Gazoletal2018ForestResilience].

The tree responses to drought are site-dependent [@Babstetal2013SiteSpecies], particularly for rear-edge populations [@CavinJump2017HighestDrought; @DoradoLinanetal2017LargescaleAtmospheric]. Greenness and tree growth were more affected by drought events in drier northern populations than in wetter southern oak populations of Sierra Nevada. The northern site showed higher browning intensity than the southern sites during the 2005 drought event, and stronger correlations of tree-growth with SPEI (hydrological year and summer) in the northern site can be interpreted as higher sensitivity to drought at drier sites [@GeaIzquierdoCanellas2014LocalClimate]. Greenness was less sensitive to drought than tree growth, particularly for drier sites. These findings agree with previous works showing tree growth to be a more sensitive metric of forest resilience than is net primary productivity [*e.g.* @Babstetal2013SiteSpecies; @Coulthardetal2017TreeGrowth; @Gazoletal2018ForestResilience; @PenaGallardoetal2018DroughtSensitiveness], suggesting that the growth reduction could be mediated by sink more than by source limitations [@Korner2013GrowthControls; @Fatichietal2014MovingPhotosynthesis]. Tree-ring records complement remote-sensing data in longer time scales by reflecting tree-growth anomalies induced by climate or disturbance over decades to centuries [@Babstetal2017ImprovedTreering] and provide an accurate measure of growth responses to droughts [@Bhuyanetal2017DifferentResponses; @Gazoletal2018ForestResilience]. On the other hand, trees at CA-High registered higher BAI than those located at lower elevations (CA-Low and SJ; Figure 4). This shows the high variability in the response to climate exhibited along a narrow gradient, which is especially noteworthy for southern sites, as these lie close to each other and both are considered to constitute the rear edge for the species.

As with many other forest species under Mediterranean climates, moisture availability is generally the most limiting factor driving radial growth of *Q. pyrenaica* along its distribution range in the Iberian Peninsula [@GeaIzquierdoCanellas2014LocalClimate]. Thus, our results are consistent with those of previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @Tessieretal1994DeciduousQuercus; @DiFilippoetal2010ClimateChange; @GeaIzquierdoetal2011TreeringsReflect; @GarciaGonzalezSoutoHerrero2017EarlywoodVessel]. A positive effect of moisture availability and negative of temperature expressing a limiting effect of high vapor pressure deficit and potential evapotranspiration can be expected at drought-limited rear-edges. Yet, at the rear-edge, growth of some tree species (*e.g.* *Abies alba*) have been shown to be more sensitive to moisture-related variables [@MartinezSanchoGutierrezMerino2019EvidenceThat], others species were more sensitive to temperatures [*e.g.* *Pinus sylvestris*, @Herreroetal2013VaryingClimate], and growth of other species responded simultaneously to both temperature and moisture related variables [*e.g.* *Fagus sylvatica*, @DoradoLinanetal2017CoexistenceMediterraneanTemperate; @DoradoLinanetal2017ClimateThreats; *Pinus nigra* subsp. *salzmanii*, @SanchezSalgueroetal2012DroughtMain]. This diversity in the response of tree species to precipitation and temperature suggests that vulnerability to climate change is not consistently expressed within the rear edge, therefore evidencing that geographically marginal forests are not necessarily climatically or ecologically marginal [see @DoradoLinanetal2019GeographicalAdaptation and references therein].

### Relict oaks show high resilience to drought at different spatio-temporal scales: do the geographical an ecological rear-edges match?
Despite the severe drought events in recent decades (Table S3), we found a positive trend for vegetation greenness of *Q. pyrenaica* for the last 16 years. This is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather wet decade in the 2000s after a dry decade in the 1990s [@PerezLuqueetal2015OntologicalSystem]. For tree growth, positive trends also appeared in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4a). Similar long-term trends have been described for this species along its distribution range only at high-elevation wet and cold sites [@GeaIzquierdoCanellas2014LocalClimate]. This could be related to a non-linear positive effect of warming for species at cold-limited high-elevation sites [@Salzeretal2009RecentUnprecedented; @GeaIzquierdoCanellas2014LocalClimate]. Importantly, for rear edges threatened by climate change, negative growth trends were expected, as shown for some temperate and Mediterranean species [@SanchezSalgueroetal2012DroughtMain; @Camareroetal2015NotEarly; @DoradoLinanetal2017CoexistenceMediterraneanTemperate].

@@

The high drought-resilience values reported here, coinciding with high values of genetic resilience for those forests on Sierra Nevada reported elsewhere [@ValbuenaCarabanaGil2013GeneticResilience; @ValbuenaCarabanaGil2017CentenaryCoppicing] appear to indicate the strong local adaptation of this oak species. Our findings agree with those of studies showing that the assumed higher vulnerability of dry edges does not necessarily hold [*e.g.* @CavinJump2017HighestDrought]. In our case, this can be explained by the fact that the current geographical rear-edge does not match with the potential ecologically rear-edge for the species because this has been determined mostly by human use. @MartinezVilalta2018RearWindow pointed out the importance of local adaptation and plasticity, and also of local environmental factors on the vulnerability shown by rear-edge populations. Our results highlight the ample small-scale variability at the ecological boundary and need to better define and delimit the rear-edges. All this, together with the characteristic high resprouting ability of the species, show the long-term persistence of these populations  [@BellinghamSparrow2000ResproutingLife]. It should be mentioned that we studied only adult individuals established decades or centuries ago, meaning that it needs to be assessed whether the high resilience observed is expressed at the species level or only in adult trees. The rear-edge might be different for different ontogenic stages. It is important to assess whether there is vulnerability if we analyzed seedling regeneration and recruitment, as in other Mediterranean species at their xeric limit [@Castroetal2004SeedlingEstablishment; @VilaCabreraetal2011StructuralClimatic; @GeaIzquierdoetal2015ThisEnd].

>>>>

The response of species to changing environments (*e.g.* distributional shifts) can be determined largely by population responses at range margins [@HampePetit2005ConservingBiodiversity]. Peripheral populations are usually considered more vulnerable compared with populations at the center of a species' range [*i.e.* center-periphery hypothesis; @SagarinGaines2002AbundantCentre; @Pirononetal2017GeographicVariation]. Geographically marginal populations have often been assumed to represent ecologically marginal populations. This means lower performance, higher vulnerability, and thus higher risk of extinction than for populations at the core of the species' range [@Rehmetal2015LosingYour; @Pirononetal2017GeographicVariation; @VilaCabreraetal2019RefiningPredictions]. Nonetheless, recent reviews report that species- and population-specific responses do not always support this hypothesis [@Sextonetal2009EvolutionEcology; @Abelietal2014EffectsMarginality; @Oldfatheretal2020RangeEdges]. This is partly because a rear-edge is a multidimensional concept including an ecological (*i.e.* climatic and edaphic), a geographical, and a genetic component [@VilaCabreraetal2019RefiningPredictions], but also an anthropogenic dimension (*i.e.* land-use). In this respect, to fully understand changes in distribution and abundance of species as a consequence of global change, it is crucial to identify and understand mismatches between the geographical and the ecological rear edges [@VilaCabreraJump2019GreaterGrowth].

Limits of species distribution are strongly determined by climatic factors and biotic interactions [@Gaston2009GeographicRange; @Sextonetal2009EvolutionEcology]. Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and signs already indicate that more intense and longer droughts are altering forest dynamics [@Allenetal2010GlobalOverview]. Drought frequency and severity have increased in recent decades, with a trend towards drier summers, particularly for Southern Europe [@VicenteSerranoetal2014EvidenceIncreasing; @Staggeetal2017ObservedDrought]. In this climatic-change context, population loss and range retractions are expected in boreal, temperate, and Mediterranean species at the lowest latitudes and elevations, as well as in drought-prone areas of a species' distribution, *i.e.* the rear edge. The rear-edge populations are likely to be more sensitive to minor climatic and microtopographic variations and therefore the effects of droughts are expected to be particularly noteworthy [@HampePetit2005ConservingBiodiversity; @VilaCabreraetal2019RefiningPredictions].

It is often overlooked that human activity constitutes a driver of change as powerful or even more powerful as natural drivers, *i.e.* natural variation in climate, particularly for regions with long land-use history such as the Mediterranean Region [*e.g.* @NavarroGonzalezetal2013WeightLanduse; @DoblasMirandaetal2017ReviewCombination]. In these areas, the susceptibility and response of ecosystems to climate change are conditioned by legacies of historical land-use activity [*e.g.* @Munteanuetal2015Legacies19th; @Mausolfetal2018LegacyEffects]. The past land-use legacies interact with recent human-caused climate disturbances and may confound their interpretation [@Fosteretal2003ImportanceLanduse]. For example, recent works showed that a quarter of current forests in the Iberian Peninsula, are growing on former agricultural and grazing land abandoned after the 1950s [@VilaCabreraetal2017NewForests]. Consequently, anthropogenic habitat modification and its legacies represent a critical dimension of marginality as they may intensify, confound or delay climate-driven population decline at rear edges [@VilaCabreraJump2019GreaterGrowth; @SanchezdeDiosetal2020FagusSylvatica]. In this context, our work seeks to identify the impacts and responses to natural (*e.g.* severe drought) and human disturbances (*e.g.* logging) on oak forests at their southern geographical range. A historical perspective should help us to interpret the responses of ecosystems to disturbances [@Fosteretal2003ImportanceLanduse], particularly regarding marginal rear-edge populations [@VilaCabreraetal2019RefiningPredictions].

The assessment of resilience to climate and human disturbances provides critical information concerning the capacity of forests to maintain their structure and render valuable ecosystem services. Resilience is the capacity of an ecosystem to persist and maintain its state and functions in the face of disturbance [@Holling1973ResilienceStability; @Hodgsonetal2015WhatYou]. @Lloretetal2011ComponentsTree proposed an approach, which decomposes resilience into three components: resistance to drought, recovery after drought and resilience. This resilience is determined by the forest's ability to mitigate the disturbance (resistance) and the capacity to recover from the impact (recovery) [@IngrischBahn2018ComparableQuantification]. This conceptual approach has recently become widely used to assess forest resilience, because it allows a simple, yet highly efficient assessment of short-term responses of trees to drought. Nevertheless, not exempt from criticism, this approach needs to be applied carefully to avoid potential bias at different levels [@Schwarzetal2020QuantifyingGrowth]. In this sense we assessed forest resilience both over the short-term to several recent extreme drought episodes, as well as over the long-term to climate change (*i.e.* warming on the last few decades), using two different proxies to characterize resilience. Dendroecological estimates of growth (*i.e.* tree-ring width) are commonly used proxies to characterize tree vitality and have commonly been used to study growth changes in response to drought at the individual tree level [@Fritts1976TreeRings; @Dobbertin2005TreeGrowth]. Remote sensing can be used to analyze the impact of drought on ecosystems at the stand level [*e.g.* @Zhangetal2013MonitoringEstimating]. Tree-ring records complement remote-sensing data with a longer time scale. Tree rings can reflect tree-growth anomalies induced by climate or other disturbance over decades to centuries [@Babstetal2017ImprovedTreering] and provide an accurate measure of growth responses to droughts [@Bhuyanetal2017DifferentResponses]. The combination of remote sensing and dendroecology has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerranoetal2013ResponseVegetation; @Coulthardetal2017TreeGrowth], and also to evaluate growth resilience to drought in several tree species [*e.g.* @Gazoletal2018ForestResilience; @PenaGallardoetal2018DroughtSensitiveness].

In the present study, we assess resilience of *Quercus pyrenaica* Willd. (Pyrenean oak) from southern relict forests at the rear edge of the species distribution, where species performance is considered to be threatened by climate change [@GeaIzquierdoetal2013GrowthProjections; @GeaIzquierdoetal2017RiskyFuture]. For this, we combined remote-sensing information and dendroecological methods to evaluate the impact of drought both on canopy greenness (as a proxy for primary growth) and radial tree growth (as a proxy for secondary growth). For the analysis of forest resilience to climate, we took into account the land-use history of these transformed forests, thoroughly reviewing historical documents to reconstruct forest history at the study sites, and analyzing how anthropogenic drivers have shaped the current forest structure. Based on this analysis, we developed a rationale that integrates the ecological and anthropogenic components of marginality to determine the regional and local scale mechanisms shaping the probability of persistence (or extinction) of rear-edge oak populations. Our main hypothesis is that range edge stands will show low resilience to extreme droughts, but that the vulnerability to drought will be reduced quickly across a fine-scale topographic gradient of decreasing aridity. To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests at their present geographical rear edge; (*ii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iii*) reviewed historical documents to reconstruct forest-management history and to infer how it impacted tree growth and stand dynamics over time; (*iv*) and examined differences in the resilience metrics between populations under contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale.

@@

The forests of this species reach their southernmost European limit in Andalusian mountains such as Sierra Nevada (37°N, 3°W), a high-mountain range with elevations of up to 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought and increasing aridity with decreasing altitude, and marked variability in annual rainfall according to elevation and aspect. Sierra Nevada is considered a glacial refuge for deciduous *Quercus* species [@Olaldeetal2002WhiteOaks]. Eight Pyrenean oak patches (2400 ha) have been identified in this mountain range (Figure 1), from 1100 to 2000 m *a.s.l.* and often associated with major river valleys. Today, *Quercus pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@HampePetit2005ConservingBiodiversity]. They are the richest forest formation in vascular plant species of Sierra Nevada, containing several endemic and endangered plant species [@Loriteetal2008PhytosociologicalReview]. They also harbor high levels of intraspecific genetic diversity [@ValbuenaCarabanaGil2013GeneticResilience]. These relict forests have undergone intensive human use throughout history [@CamachoOlmedoetal2002DinamicaEvolutiva]. Furthermore, the conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer from climate change, potentially reducing its suitable habitats in the near future [@GeaIzquierdoetal2013GrowthProjections; @GeaIzquierdoetal2017RiskyFuture].

@@

The severe drought events since 1901 were identified using SPEI-12. They were characterized in terms of duration, severity and intensity [see @Spinonietal2015EuropeanDrought]. In a first step, the impact of drought in greenness and growth was explored using the $EVI_{sa}$ and the mean RWI site chronologies (Figure 2). Additionally the relationships between climatic variables and tree-growth variables (RWI and BAI site chronologies) were assessed using bootstrapped Pearson's correlations estimated using treeclim [@ZangBiondi2015TreeclimPackage]. The non-climatic disturbance impacts on tree-growth were evaluated using site disturbance chronologies (built using growth changes on tree growth).

@@

## Results
### Temporal trends in vegetation greenness
The analysis of temporal trends in greenness showed that 78.9% of the EVI pixels followed a positive trend for the 2000-2016 period. The lowest values of EVI standardized anomalies for the study period were recorded during the 2005 drought, and the minimum EVI values were expressed in the northern (dry) population (Figure 3a). A "browning" episode ($\mathrm{EVI_{sa}} < -1$) was found during this drought event, whereas no changes in greenness in response to the 2012 drought were detected (Figure 3b).

### Analysis of radial-growth trends and disturbances 
The trees of the southern population were older than those from the northern one. In addition, trees from the southern population at high elevation were taller and their growth was significantly greater than that of trees from the other two sites. Stand competition measured as plot basal area was greatest in CA-High (Table 1, Figure 4a). The growth and height of trees from the northern and the low-elevation southern population were similar (Figures 4a and S3a). Only trees from the southern sites (*i.e.* the wetter exposure) showed significant positive growth trends since the late 1970s (Figure 4a), this trend being far more pronounced for the wetter and colder high elevation site (CA-High).

@@

### Resilience, resistance and recovery to drought events at the stand and individual-tree levels
Resilience and resistance varied in the same direction whereas recovery varied inversely to resilience and resistance. During the last two drought events, resilience metrics for greenness and tree growth significantly differed between drought events (Table S1). The 2005 drought event reduced greenness and growth more than that of 2012 (Table S2) but the metrics of resilience generally covaried in the same direction during those two drought events. For EVI, resilience and resistance values were significantly higher for 2012, the most severe event, than for 2005 (Table S2; Figure 5b); whereas recovery values were higher for 2005 than for the 2012 drought event. For BAI, the resilience, resistance and recovery values were higher for 2012 than for 2005 (Table S2, Figure 5c).

@@

## Discussion
By using a combined approach of remote-sensing information and dendroecology, we quantified the growth of adult trees and greenness (EVI) as proxies for secondary and primary growth of relict Mediterranean *Quercus pyrenaica* populations in the southern Iberian Peninsula. These relict oak populations, driven by historical land-use, have been resilient to climate change at their present rear edge. However, resistance, resilience, and forest recovery after extreme drought events were strongly influenced by mountain exposure, local environmental conditions, and management legacies. This means that the geographical and the ecological rear edges do not necessarily match and, at a small spatial scale, tree performance can vary markedly along the rear edge under climate change.

### High sensitivity and variability in the oak sensitivity to climate at the rear edge
Severe drought negatively affects both primary and secondary growth of *Q. pyrenaica* forests. This was expressed by the observed reduction in greenness and tree growth in response to the 2005 and 2012 drought events as well as by radial-growth suppression during extreme drought events [@Corcueraetal2006RadialgrowthWoodanatomical; @GeaIzquierdoCanellas2014LocalClimate]. Furthermore, the greatest reduction of tree growth was detected during the 1995 drought, a characteristic negative precipitation anomaly that caused severe and extensive damage in the Mediterranean Iberian Peninsula [@Penuelasetal2001SevereDrought; @Gazoletal2018ForestResilience].

The tree responses to drought are site-dependent [@Babstetal2013SiteSpecies], particularly for rear-edge populations [@CavinJump2017HighestDrought; @DoradoLinanetal2017LargescaleAtmospheric]. Greenness and tree growth were more affected by drought events in drier northern populations than in wetter southern oak populations of Sierra Nevada. The northern site showed higher browning intensity than did the southern sites during the 2005 drought event, and stronger correlations of tree-growth with SPEI (hydrological year and summer) at the northern site can be interpreted as higher sensitivity to drought at drier sites [@GeaIzquierdoCanellas2014LocalClimate]. Greenness was less sensitive to drought than tree growth, particularly for drier sites. These findings agree with previous works showing tree growth to be a more sensitive metric of forest resilience than is net primary productivity [*e.g.* @Babstetal2013SiteSpecies; @Coulthardetal2017TreeGrowth; @Gazoletal2018ForestResilience; @PenaGallardoetal2018DroughtSensitiveness], suggesting that the growth reduction could be mediated by sink more than by source limitations [@Korner2013GrowthControls; @Fatichietal2014MovingPhotosynthesis]. On the other hand, trees at CA-High registered higher BAI than did those located at lower elevations (CA-Low and SJ; Figure 4). This shows the high variability in the response to climate exhibited along a narrow gradient, which was especially noteworthy for southern sites, as these lie close to each other and both are considered to constitute the rear edge for the species.

As with many other forest species under Mediterranean climates, moisture availability is generally the most limiting factor driving radial growth of *Q. pyrenaica* along its distribution range in the Iberian Peninsula [@GeaIzquierdoCanellas2014LocalClimate]. Thus, our results are consistent with those of previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @Tessieretal1994DeciduousQuercus; @DiFilippoetal2010ClimateChange; @GeaIzquierdoetal2011TreeringsReflect; @GarciaGonzalezSoutoHerrero2017EarlywoodVessel]. A positive effect of moisture availability and negative impact of temperature expressing a limiting effect of high vapor-pressure deficit and potential evapotranspiration can be expected at drought-limited rear-edges. Yet, at the rear edge, the growth of some tree species (*e.g.* *Abies alba*) has been shown to be more sensitive to moisture-related variables [@MartinezSanchoGutierrezMerino2019EvidenceThat], while others species were more sensitive to temperatures [*e.g.* *Pinus sylvestris*, @Herreroetal2013VaryingClimate], and still other species responded simultaneously to both temperature and moisture-related variables [*e.g.* *Fagus sylvatica*, @DoradoLinanetal2017CoexistenceMediterraneanTemperate; @DoradoLinanetal2017ClimateThreats; *Pinus nigra* subsp. *salzmanii*, @SanchezSalgueroetal2012DroughtMain]. This diversity in the response of tree species to precipitation and temperature suggests that vulnerability to climate change is not consistently expressed within the rear edge, therefore evidencing that geographically marginal forests are not necessarily climatically or ecologically marginal [see @DoradoLinanetal2019GeographicalAdaptation and references therein].

### Relict oaks show high resilience to drought at different spatio-temporal scales: do the geographical and ecological rear-edges match?
Despite the severe drought events in recent decades (Table S3), we found a positive trend for vegetation greenness of *Q. pyrenaica* for the last 16 years. This is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather wet decade in the 2000s after a dry decade in the 1990s [@PerezLuqueetal2015OntologicalSystem]. For tree growth, positive trends also appeared in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4a). Similar long-term trends have been described for this species along its distribution range only at high-elevation wet and cold sites [@GeaIzquierdoCanellas2014LocalClimate]. This could be related to a non-linear positive effect of warming for species at cold-limited high-elevation sites [@Salzeretal2009RecentUnprecedented; @GeaIzquierdoCanellas2014LocalClimate]. Importantly, for rear edges threatened by climate change, negative growth trends were expected, as shown for some temperate and Mediterranean species [@SanchezSalgueroetal2012DroughtMain; @Camareroetal2015NotEarly; @DoradoLinanetal2017CoexistenceMediterraneanTemperate].

@@

The high drought-resilience values reported here, in addition to the potential role of local adaptation [*i.e.* high values of genetic resilience for oak forests on Sierra Nevada; @ValbuenaCarabanaGil2013GeneticResilience; @ValbuenaCarabanaGil2017CentenaryCoppicing], suggest that land-use also has a key role to determine tree resilience to drought and the range edge of species. Our findings agree with those of studies showing that the assumed higher vulnerability of current geographical dry edges does not necessarily hold [*e.g.* @CavinJump2017HighestDrought]. In our case, this can be explained by the fact that the current geographical rear-edge does not match with the potential ecological rear edge for the species because this has been modified and determined mostly by human use. @MartinezVilalta2018RearWindow pointed out the importance of local adaptation and plasticity, and also of local environmental factors on the vulnerability shown by rear-edge populations. Our results highlight the ample small-scale variability at the ecological boundary and thus the rear edges need to be more clearly defined and delineated. All the above points, together with the characteristic high resprouting ability of the species, show the long-term persistence of these populations  [@BellinghamSparrow2000ResproutingLife]. It should be mentioned that we studied only adult individuals established decades or centuries ago, meaning that it needs to be assessed whether the high resilience found is expressed at the species level (*i.e.* also including regeneration) or only in adult trees. The rear-edge might differ for different ontogenic stages. It is important to assess whether seedling regeneration and recruitment are vulnerable, as in other Mediterranean species at some locations including their xeric limit [@Castroetal2004SeedlingEstablishment; @VilaCabreraetal2011StructuralClimatic; @GeaIzquierdoetal2015ThisEnd].
